# Food bank accessibility

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
```

::: callout-note
**Aim: Identify areas (in and around Oxford/shire) with low access and high demand for food banks**

-   Convert food bank addresses to point locations
-   Create maps/visualisations of locations of food banks (in the UK and Oxfordshire)
-   Calculate accessibility for walking, driving, and public transport using r5r package
-   Identify areas with low accessibility to food banks, but a perceived high need (based on a couple of factors)
:::

::: callout-note
**Optional reading**

-   Pereira, R. et al., [r5r: Rapid Realistic Routing on Multimodal Transport Networks with R5 in R (2021)](https://doi.org/10.32866/001c.21262)
-   Allen, J. and Farber, S. [Changes in transit accessibility to food banks in Toronto during Covid-19 (2021)](https://doi.org/10.32866/001c.24072)
-   Higgins, C.D. et al. [Changes in accessibility to emergency and community food services during Covid-19 and implications for low income populations in Hamilton, Ontario (2021)](https://pubmed.ncbi.nlm.nih.gov/34655939/)
:::

## Convert food bank addresses to point locations

Firstly, load the R packages needed for the practical. You may need to install some of these for the first time.

```{r packages}
#Load relevant packages
library(dplyr)
library(tidyverse)
library(janitor)
library(RColorBrewer)
library(ggplot2) #for graphs
library(tmap) #for maps
library(tmaptools) #for maps
library(leaflet) #for maps
library(reader)
library(readxl)
library(data.table)
library(knitr)
library(sp)
library(sf)
library(spdep)
library(spatstat)
library(PostcodesioR) #for postcode lat and long
library(mosaic) #for favstats function
library(geodist) #for distances
library(rje) #for row mins
library(osmdata) #for osm
library(osmextract) # for osm
library(osrm) #for osm
library(r5r) #for public transport
library(tidygeocoder)
library(tidytransit) #for public transport
```

### Load the spatial data

We need to load the spatial data for different spatial scales. This polygons data will be used for creating maps and calculating accessibility to food banks centres. These were downloaded from the [Open Geography Portal](https://geoportal.statistics.gov.uk) (from the Office for National Statistics).

-   Lower super output area (LSOA) scale
-   Local authority scale
-   Entire country scale

There are 35,672 LSOAs (for England and Wales)

```{r load LSOAs}
#2021 shape file for LSOAs
LSOA_2021 <- st_read("prac5_data/LSOA_2021/LSOA_(Dec_2021)_Boundaries_Generalised_Clipped_EW_(BGC).shp") %>%
  st_transform(., 27700) %>%
  clean_names() 
```

There are 374 local authorities (for the UK)

```{r load Local Authority districts}
#2022 shape file for LAs
LA_2022 <- st_read("prac5_data/LA_2022/LAD_MAY_2022_UK_BFE_V3.shp") %>%
  st_transform(., 27700) %>%
  clean_names() 
```

```{r load country borders}
#2023 shape file for UK
UK_border <- st_read("prac5_data/UK_2023/CTRY_DEC_2023_UK_BUC.shp") %>%
  st_transform(., 27700) %>%
  clean_names() 
```

### Data on Trussell food bank centres

Next, we load the Trussell food bank data with addresses (this data was provided by Trussell in February 2023 so some centres may have closed in the last two years or new centres may have opened).

```{r locations of Trussell food banks}
#locations of food banks from TT
TT_foodbanks <- read_excel("prac5_data/Food banks Feb23.xlsx") %>%
  clean_names()
```

In order to derive point locations from postcodes, we need to clean the food bank postcodes so they are in a uniform and readable form. At this stage, we also remove food bank centres that are not currently open or are warehouses or offices (where people cannot collect food from).

```{r clean postcodes and remove certain records}
#removing white space from postcodes
TT_foodbanks$postcode <- gsub("[[:space:]]", "", TT_foodbanks$postcode)

#filter out only distribution centres
TT_foodbanks <- TT_foodbanks %>%
  filter(type == "Distribution Centre") %>% #only want distribution centres and not warehouses or offices
  filter(operating_model != "Temporarily Closed") #remove centres that are currently closed
```

Next, we derive the point locations from postcodes. Unfortunately, the approach below only allows 100 locations at a time to be converte and the lat/lon will be a centroid of the postcode not the actual location of the distribution centre.

The function `lapply` used in these code chunks applies a function over a list or vector. We are applying it to a list of postcodes.

```{r locations of TT food banks 1, include = TRUE}
TT_foodbanks1 <- TT_foodbanks[1:100, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks1$postcode))

#foodbank locations
#lat and long from postcodes using bulk postcode look up function - limited to 100
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up - extract 2nd element from the list
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points1 <-
  # iterate over each element and extract the following into a dataframe
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points1 <- 
  st_as_sf(TT_foodbank_location_points1,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 2, include = TRUE}
TT_foodbanks2 <- TT_foodbanks[101:200, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks2$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points2 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points2 <- 
  st_as_sf(TT_foodbank_location_points2,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 3, include = TRUE}
TT_foodbanks3 <- TT_foodbanks[201:300, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks3$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points3 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points3 <- 
  st_as_sf(TT_foodbank_location_points3,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 4, include = TRUE}
TT_foodbanks4 <- TT_foodbanks[301:400, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks4$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points4 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points4 <- 
  st_as_sf(TT_foodbank_location_points4,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 5, include=TRUE}
TT_foodbanks5 <- TT_foodbanks[401:500, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks5$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points5 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points5 <- 
  st_as_sf(TT_foodbank_location_points5,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 6, include=TRUE}
TT_foodbanks6 <- TT_foodbanks[501:600, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks1$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points6 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points6 <- 
  st_as_sf(TT_foodbank_location_points6,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 7, include=TRUE}
TT_foodbanks7 <- TT_foodbanks[601:700, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks7$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points7 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points7 <- 
  st_as_sf(TT_foodbank_location_points7,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 8, include=TRUE}
TT_foodbanks8 <- TT_foodbanks[701:800, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks8$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points8 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points8 <- 
  st_as_sf(TT_foodbank_location_points8,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 9, include=TRUE}
TT_foodbanks9 <- TT_foodbanks[801:900, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks9$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points9 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points9 <- 
  st_as_sf(TT_foodbank_location_points9,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 10, include=TRUE}
TT_foodbanks10 <- TT_foodbanks[901:1000, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks10$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points10 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points10 <- 
  st_as_sf(TT_foodbank_location_points10,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 11, include=TRUE}
TT_foodbanks11 <- TT_foodbanks[1001:1100, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks11$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points11 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points11 <- 
  st_as_sf(TT_foodbank_location_points11,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 12, include=TRUE}
TT_foodbanks12 <- TT_foodbanks[1101:1200, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks12$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points12 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points12 <- 
  st_as_sf(TT_foodbank_location_points12,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 13, include=TRUE}
TT_foodbanks13 <- TT_foodbanks[1201:1300, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks13$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points13 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points13 <- 
  st_as_sf(TT_foodbank_location_points13,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 14, include=TRUE}
TT_foodbanks14 <- TT_foodbanks[1301:1400, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks14$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points14 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points14 <- 
  st_as_sf(TT_foodbank_location_points14,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r locations of TT food banks 15, include=TRUE}
TT_foodbanks15 <- TT_foodbanks[1401:1425, ]

#make postcodes into a list
list <- as.list(unique(TT_foodbanks15$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

TT_foodbank_location_points15 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
TT_foodbank_location_points15 <- 
  st_as_sf(TT_foodbank_location_points15,
           coords = c("longitude", "latitude"), crs = 4326)
```

After all the Trussell centres have had their point location derived, we bind them together in one data frame.

```{r bind all food bank locations together}
TT_foodbank_location_points <- rbind(TT_foodbank_location_points1, TT_foodbank_location_points2,
                                     TT_foodbank_location_points3, TT_foodbank_location_points4, 
                                     TT_foodbank_location_points5, TT_foodbank_location_points6,
                                     TT_foodbank_location_points7, TT_foodbank_location_points8,
                                     TT_foodbank_location_points9, TT_foodbank_location_points10,
                                     TT_foodbank_location_points11, TT_foodbank_location_points12,
                                     TT_foodbank_location_points13, TT_foodbank_location_points14,
                                     TT_foodbank_location_points15)
```

```{r how many food bank centres have been lost?}
nrow(TT_foodbanks) - nrow(TT_foodbank_location_points)
```

This process has lost 42 food bank centres as postcode could not be recognised/matched with a longitude and latitude.

We need to join the point location to the other information on the food bank centres like the centre's name and the operating model etc.

```{r editing Trussell location df}
#removing white space from postcodes in order to merge with more information
TT_foodbank_location_points$postcode <- gsub("[[:space:]]", "", TT_foodbank_location_points$postcode)

#joining point data with other information on food bank centres e.g. name and operating model
TT_foodbank_location_points <- TT_foodbank_location_points %>%
  merge(.,
        TT_foodbanks, 
        by.x = "postcode",
        by.y = "postcode",
        all.x = TRUE)

#select specific columns
TT_foodbank_location_points <- TT_foodbank_location_points %>%
  distinct(foodbank_centre_number, postcode, .keep_all = TRUE) %>%
  dplyr::select(food_bank_name, distribution_centre_name, foodbank_centre_number, postcode, operating_model, number_of_distribution_centres_final)
```

### Create a map of Trussell food bank locations in the UK

Now we have the point location of Trussell centres, we can use the tmap package to visualise *where* the Trussell food bank centres are across the UK by making a simple map.

```{r where are the food banks}
TT_foodbank_location_points <- TT_foodbank_location_points %>%
  st_transform(., 27700) 

#spatially subset (select) the point locations to be only within the shape file area - excludes food banks outside the boundary
# defaults to st_intersects
TT_foodbank_location_points <- TT_foodbank_location_points[UK_border,] 

#plot map
Plain_map <- 
  tm_shape(UK_border) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(TT_foodbank_location_points) +
  tm_symbols(col = "forestgreen", size = 0.03) +
  tm_scale_bar(position = c("left", "top")) + 
  tm_compass(position = c("right", "top")) + 
  tm_layout(main.title = "Trussell food bank centres", 
            main.title.position = "centre",
            main.title.size = 0.7, 
            legend.outside = TRUE, frame = FALSE)

Plain_map
```

::: callout-note
**Questions**

-   What does this tell up about the spatial distribution of food banks?
-   What could be added to this map to enhance the visualisation?
:::

## Create a map of Trussell food bank locations in Oxford and Oxfordshire

Let's zoom in on Oxford and the surrounding area.

```{r filter LAs for Oxford and Oxfordshire}
#use string detect function to find local authorities with these words in their names
Oxford_LAs <- LA_2022 %>%
  filter(str_detect(lad22nm, "Oxford") | 
           str_detect(lad22nm, "White Horse") | 
           str_detect(lad22nm, "Cherwell")) 
```

```{r filter LAs for Oxford}
#use string detect function to isolate the border of the city of Oxford
Oxford_City_LAs <- LA_2022 %>%
  filter(str_detect(lad22cd, "E07000178")) 
```

```{r map of Trussell food bank centres in Oxfordshire}
#find food bank centres from within the border we have created
Oxford_TT_foodbanks <- TT_foodbank_location_points[Oxford_LAs,] 

Oxford_map <- 
  tm_shape(Oxford_LAs) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + #bolder border around the City of Oxford
  tm_shape(Oxford_TT_foodbanks) +
  tm_symbols(col = "deepskyblue", size = 0.1) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
  tm_layout(main.title = "Trussell centres in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9, 
            legend.outside = TRUE, frame = FALSE)

Oxford_map
```

```{r interactive map of Trussell food banks in Oxfordshire}
tmap_mode("view")

#plot map
Oxford_map <- 
  tm_shape(Oxford_LAs) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + #bolder border around the City of Oxford
  tm_shape(Oxford_TT_foodbanks) +
  tm_symbols(col = "deepskyblue", size = 0.1) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
    tm_layout(main.title = "Trussell centres in/around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9, 
            legend.outside = TRUE, frame = FALSE)

Oxford_map
```

We see only seven food bank centres across the region, and no food bank centres in the City of Oxford itself, so we look at other sources of data to see if this is true.

### Give Food food bank data

Give Food is an open access database with information on food banks in the UK. Next, load the food bank data with addresses (from [Give Food](https://github.com/givefood/data)). However, these are food banks and not distribution centres.

```{r locations of food banks (not centres)}
#data from give food
give_food_foodbanks <- read_csv("prac5_data/foodbanks.csv") %>%
  clean_names()
```

```{r clean postcodes and remove certain records for give food}
#removing white space from postcodes
give_food_foodbanks$postcode <- gsub("[[:space:]]", "", give_food_foodbanks$postcode)

#filter out only distribution centres
give_food_foodbanks <- give_food_foodbanks %>%
  filter(closed != "TRUE") #remove centres that are currently closed
```

```{r detect centres in Oxford and Oxfordshire}
#create 2 versions
Oxford_give_food_foodbanks <- give_food_foodbanks %>%
  filter(str_detect(postcode, "OX")) 

Oxford_give_food_foodbanks2 <- give_food_foodbanks %>%
  filter(str_detect(postcode, "OX")) 
```

```{r locations of give food food banks, echo=FALSE, include=FALSE}
#make postcodes into a list
list <- as.list(unique(Oxford_give_food_foodbanks2$postcode))

#foodbank locations
#lat and long from postcodes
pc_list <- list(postcodes = list)
bulk_lookup_result <- bulk_postcode_lookup(pc_list)
str(bulk_lookup_result[1])

#postcode look up
bulk_list <- lapply(bulk_lookup_result, "[[", 2)

Oxford_give_food_foodbanks2 <-
  map_dfr(bulk_list,
          `[`,
          c("postcode", "longitude", "latitude"))

#as spatial points
Oxford_give_food_foodbanks2 <- 
  st_as_sf(Oxford_give_food_foodbanks2,
           coords = c("longitude", "latitude"), crs = 4326)
```

```{r join point data to original df}
Oxford_give_food_foodbanks2$postcode <- gsub("[[:space:]]", "", Oxford_give_food_foodbanks2$postcode)

#create spatial object
Oxford_give_food_foodbanks2 <- Oxford_give_food_foodbanks2 %>%
  st_transform(., 27700) 

#select only specific variables needed
Oxford_give_food_foodbanks <- Oxford_give_food_foodbanks %>%
  select(name, postcode, district, network)

Oxford_give_food_foodbanks <- Oxford_give_food_foodbanks2 %>%
  merge(.,
        Oxford_give_food_foodbanks, 
        by.x = "postcode",
        by.y = "postcode",
        all.x = TRUE)
```

```{r map other food banks in Oxfordshire}
#create spatial object
Oxford_give_food_foodbanks <- Oxford_give_food_foodbanks %>%
  st_transform(., 27700) 

tmap_mode("plot")

#plot map
Oxford_map <- 
  tm_shape(Oxford_LAs) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + #bolder border around the City of Oxford
  tm_shape(Oxford_give_food_foodbanks) +
  tm_symbols(col = "red", size = 0.1) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
  tm_layout(main.title = "Other food banks in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9, 
            legend.outside = TRUE, frame = FALSE)

Oxford_map
```

```{r interactive map other food banks in Oxfordshire}
tmap_mode("view")

#plot map
Oxford_map <- 
  tm_shape(Oxford_LAs) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + #bolder border around the City of Oxford
  tm_shape(Oxford_give_food_foodbanks) +
  tm_symbols(col = "red", size = 0.1) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
    tm_layout(main.title = "Other food banks in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9, 
            legend.outside = TRUE, frame = FALSE)

Oxford_map
```

If you hover over the food bank centres on the interactive map, information on the centre that is contained in the data frame will pop up in a box.

Next we map the same data, but spilt the centres by network type.

```{r filter data by food network network and map it}
trussell_Oxford <- Oxford_give_food_foodbanks %>%
  filter(network == "Trussell")

IFAN_Oxford <- Oxford_give_food_foodbanks %>%
  filter(network == "IFAN")

independent_Oxford <- Oxford_give_food_foodbanks %>%
  filter(network == "Independent")

#change mode from interactive to static
tmap_mode("plot")

#plot map
Oxford_map <- 
  tm_shape(Oxford_LAs) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + #bolder border around the City of Oxford
  tm_shape(trussell_Oxford) +
    tm_symbols(col = "purple", size = 0.15) +
  tm_shape(IFAN_Oxford) +
    tm_symbols(col = "orange", size = 0.15) +
  tm_shape(independent_Oxford) +
    tm_symbols(col = "red", size = 0.15) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Other food banks in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) +
  tm_add_legend('symbol', 
	col = c("purple", "orange", "red"),
	border.col = "black",
	size = 1,
	labels = c('Trussell',
	           'IFAN',
	           'Independent'))

Oxford_map
```

Give Food is an open access database with information on food banks in the UK. Next, load the food bank data with addresses (from Give Food). However, these are **food banks** (and could under represent the sites) and not distribution centres.

```{r map all food banks}
#change mode from interactive to static
tmap_mode("plot")

#plot map
Oxford_map <- 
  tm_shape(Oxford_LAs) +
  tm_fill(col = "antiquewhite", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_TT_foodbanks) +
    tm_symbols(col = "purple", size = 0.15) +
  tm_shape(trussell_Oxford) +
    tm_symbols(col = "purple", size = 0.15) +
  tm_shape(IFAN_Oxford) +
    tm_symbols(col = "orange", size = 0.15) +
  tm_shape(independent_Oxford) +
    tm_symbols(col = "red", size = 0.15) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Other food banks in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) +
  tm_add_legend('symbol', 
	col = c("purple", "orange", "red"),
	border.col = "black",
	size = 1,
	labels = c('Trussell',
	           'IFAN',
	           'Independent'))

Oxford_map
```

## Calculate accessibility for walking, driving, and public transport using r5r package

r5r is an R package for rapid realistic routing on multi-modal transport networks (walk, bike, public transport and car).

To use r5r, we will need:

-   A road network data set from OpenStreetMap in .pbf format (mandatory)
-   A public transport feed in GTFS.zip format (optional)
-   A raster file of Digital Elevation Model data in .tif format (optional)

Link to R package page: <https://ipeagit.github.io/r5r/>.

### GTFS files

The subsections below detail

#### Bus, Trams and Tubes

There are a couple of websites to get a UK GTFS feed:

-   [Mobility Database](https://mobilitydatabase.org)
-   [Traveline National Dataset (TNDS)](https://www.travelinedata.org.uk/traveline-open-data/traveline-national-dataset/)

At the moment the TNDS is the best option - as it is divided into regions (and actually works). **Note** this is just for Bus, Trams and Tubes.

To get the data:

-   Register for free
-   Download a FTP software, e.g. [Core FTP](https://www.coreftp.com/download.html)
-   Follow the instructions in the TNDS email and download data for the South East.

This data is in a TransXchange format that we must convert to a GTFS through the UK2GTFS package.

```{r,eval=FALSE}
remotes::install_github("ITSleeds/UK2GTFS")
```

The `UK2GTFS` package requires National Public Transport Access Nodes (NapTAN). It should download it automatically, but if it doesn't download it from [gov.uk](https://beta-naptan.dft.gov.uk/Download/National) and place it in a folder called temp_naptan in the root of you project (i.e. in the same places as the .RProj file)

```{r, eval=FALSE}
library(UK2GTFS)

# you might not need this
naptan_local <- read_csv("temp_naptan/Stops.csv", show_col_types = FALSE) %>%
  rename(
    stop_id = ATCOCode,
    stop_name = CommonName,
    stop_lat = Latitude,
    stop_lon = Longitude,
    stop_type = StopType
  )

# %>%
#    filter(!is.na(stop_id),
#          !is.na(stop_name),
#          !is.na(stop_lat),
#          !is.na(stop_lon))




path_in <- "prac5_data/SE.zip"
gtfs <- transxchange2gtfs(path_in = path_in,
                          # remove this line if it is not local
                          naptan = naptan_local,
                          ncores = 5)

Oxford_LAs_wgs84 <- Oxford_LAs %>%
st_transform(., 4326)
  
gtfs_oxford <- gtfs_clip(gtfs, Oxford_LAs_wgs84)
```

Save the GTFS file

```{r, eval=FALSE}
gtfs_write(gtfs, 
           folder = "prac5_data/r5r/",
           name = "gtfs_SE")
```

If errors appear with the GTFS file, either now or later, try comparing to [best practice standards](https://gtfs-validator.mobilitydata.org/) or the code below

```{r, eval=FALSE}
# merge error
gtfs <- gtfs_merge(gtfs, force = TRUE)
# clean the file
gtfs <- gtfs_clean(gtfs)
# force it to be valid
gtfs <- gtfs_force_valid(gtfs)
```

#### Trains

Rail data comes from the [National Rail Portal](https://opendata.nationalrail.co.uk/) in an ATOC CIF that we must convert to GTFS. To access this you must register for a free account. `UK2GTFS` can download the relevant file, however you must set the following in the `.Renviro` file:`NRDP_username` and `NRDP_password`. You can access this file through the `usethis` package: `usethis::edit_r_environ()`. For example:

`NRDP_username`="your username"

Restart R after making these edits.

```{r, eval=FALSE}
nrdp_timetable("prac5_data/trains/timetable.zip")
```

Convert the ATOC CIF to GTFS

```{r, eval=FALSE}
path_in <- "prac5_data/trains/timetable.zip"
gtfs <- atoc2gtfs(path_in = path_in,
          ncores = 5)


```

Save the GTFS

```{r, eval=FALSE}
gtfs_write(gtfs, 
           folder = "prac5_data/r5r/",
           name = "gtfs_rail")
```

### Centroids from LSOAs

First we need to get the point locations for LSOA centroids for Oxford and the surrounding area. You can also download the population weighted centroids for LSOAs, but we are not using those today.

Repeat the filtering process for LSOAs like we did for LAs.

```{r filter LSOAs for Oxford and Oxfordshire}
Oxford_LSOAs <- LSOA_2021 %>%
  filter(str_detect(lsoa21nm, "Oxford") | 
           str_detect(lsoa21nm, "White Horse") | 
           str_detect(lsoa21nm, "Cherwell")) 
```

```{r food bank maps for LSOA scale}
#change mode from interactive to static
tmap_mode("plot")

#plot map
Oxford_map_LSOAs <- 
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "white", alpha = 0.5) +
  tm_borders(col = "black", lwd = 0.1) +
  tm_shape(Oxford_City_LAs) +
  tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_TT_foodbanks) +
    tm_symbols(col = "purple", size = 0.15) +
  tm_shape(trussell_Oxford) +
    tm_symbols(col = "purple", size = 0.15) +
  tm_shape(IFAN_Oxford) +
    tm_symbols(col = "orange", size = 0.15) +
  tm_shape(independent_Oxford) +
    tm_symbols(col = "red", size = 0.15) +
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) + 
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Food bank centres in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) +
  tm_add_legend('symbol', 
	col = c("purple", "orange", "red"),
	border.col = "black",
	size = 1,
	labels = c('Trussell',
	           'IFAN',
	           'Independent'))

Oxford_map_LSOAs
```

By mapping food bank centres at the LSOA scale rather than the LA scale, we can assume that they are usually located in more densely populated areas like towns/villages in the area.

```{r centroids of LSOAs}
sf_use_s2(FALSE)

centroids <- st_centroid(Oxford_LSOAs) %>%
  st_transform(.,4326)
  
centroids <- centroids %>%
  mutate(long = unlist(map(centroids$geometry,1)),
           lat = unlist(map(centroids$geometry,2)))

centroids$key <- 1
centroids$geometry <- NULL

centroids <- centroids %>%
  dplyr::select(lsoa21cd, long, lat, key)
```

```{r create the origin points - LSOAs}
# load in origin locations - LSOA centroids
points_o <- centroids
points_o$lon <- as.numeric(points_o$long)
points_o$lat <- as.numeric(points_o$lat)
points_o$id <- points_o$lsoa21cd
points_o <- points_o[,c("id","lon","lat")]

#look at the data
head(points_o,10)
```
We now have a **choice to make** - what to use as a destination point, either:

* Trussell Trust data
  * We can filter distribution (access) points
  * But we have taken postcodes and assigned a lat/lon to them
  * The lat/lon will be a centroid of the postcode not the actual location of the distribution centre.
  
* Give food data
  * Provided location, so assumed to be more accurate locations
  * However, the points could be a mixture of food banks, distribution centres and even offices!
  
Here, I have used give food data.  

```{r create the destination points - centres}
Oxford_give_food_foodbanks <- Oxford_give_food_foodbanks %>%
  sf::st_transform(crs = 4326)

#unlist geometry column
Oxford_give_food_foodbanks <- Oxford_give_food_foodbanks %>%
  mutate(longitude = unlist(map(Oxford_give_food_foodbanks$geometry,1)),
           latitude = unlist(map(Oxford_give_food_foodbanks$geometry,2)))

points_d <- Oxford_give_food_foodbanks
points_d$lon <- as.numeric(points_d$longitude)
points_d$lat <- as.numeric(points_d$latitude)
points_d$id <- points_d$postcode
points_d <- points_d[,c("id","lon","lat")]


#look at the data
head(points_d,10)
```
If you look at the `ID` column it is now postcode, however might need this to be an LSOA. So we must use a [lookup table](https://open-geography-portalx-ons.hub.arcgis.com/datasets/ons::postcode-to-oa-2021-to-lsoa-to-msoa-to-lad-february-2025-best-fit-lookup-in-the-uk/about) to convert it back

```{r}
Lookup <- read_csv("prac5_data/PCD_OA21_LSOA21_MSOA21_LAD_FEB25_UK_LU.csv")

# remove all spaces in postcode
Lookup <- Lookup %>%
  mutate(pcd7 = str_replace_all(pcd7, " ", ""))

#join oxford walking time to food banks 
#select only the postcode and LSOA from lookup
# join on the postcode
points_d <- points_d %>%
  left_join(Lookup %>% select(pcd7, lsoa21cd), 
            by = c("id" = "pcd7")) %>%
  select(lsoa21cd, lon, lat, id)%>%
  # must have an id
    rename(postcode=id)%>%
  rename(id=lsoa21cd)%>%
  mutate(numeric_id=row_number())
```

The next code chunk downloads the road network for the area. However, for this practical, it is included in the data provided (as is the timetable data for public transport).

```{r download road network}
# load libraries
#library(r5r)
#https://download.geofabrik.de/europe.html

#osm_lines = oe_get("Oxfordshire", stringsAsFactors = FALSE, quiet = TRUE)
#plot(st_geometry(osm_lines))

#EW_url = oe_match("Oxfordshire")
#oe_download(
  #file_url = EW_url$url,
  #file_size = EW_url$file_size,
  #download_directory = tempdir())

#EW_url[["url"]]
```

Next we need to set up the r5r core. This sets up the road network and transit zip files in a folder. Note, once this is run once it will save the output as a `network.dat` file and load from it in future.

```{r}
# increase Java memory
options(java.parameters = "-Xmx6G")

# load libraries
library(r5r)

#path <- system.file("r5r_path", package = "r5r")

r5r_core <- setup_r5(data_path = "prac5_data/r5r.")
```

For this example, I set the maximum trip duration to 1 hour (60 minutes) and set the departure time at 2pm.

```{r public transport travel times}
# 
# gtfs_rail <- read_gtfs("prac5_data/r5r/gtfs_rail.zip")
# 
# range(gtfs_rail$calendar$start_date)
# range(gtfs_rail$calendar$end_date)
# 
# # Summarize all service periods
# gtfs_rail$calendar %>%
#   select(service_id, start_date, end_date) %>%
#   arrange(start_date)
# 
# 
# gtfs_bus <- read_gtfs("prac5_data/r5r/gtfs_SE.zip")

# 
# # Set your desired date
# dep_date <- ymd("2026-01-01")  # replace with your date
# 
# # Check calendar periods
# gtfs_bus$calendar %>%
#   mutate(start_date = ymd(start_date),
#          end_date = ymd(end_date)) %>%
#   filter(dep_date >= start_date & dep_date <= end_date)
# 
# # Check calendar_dates for exceptions
# if("calendar_dates" %in% names(gtfs_bus)) {
#   gtfs_bus$calendar_dates %>%
#     mutate(date = ymd(date)) %>%
#     filter(date == dep_date)
# }



# routing paramaters
mode <- c("TRANSIT")

max_trip_duration <- 60 # minutes

# can change to just a day (no time) and it will return min time of the entire day (to each Food bank within 60 mins)
departure_datetime <- as.POSIXct("01-03-2021 14:00:00", 
                                 format = "%d-%m-%Y %H:%M:%S",
                                 tz = "Europe/London")

# estimating the travel time
TTM_Oxford_PT <- travel_time_matrix(r5r_core,
                          origins = points_o,
                          destinations = points_d,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_trip_duration = max_trip_duration)
```

```{r walking travel times}
# routing paramaters
mode <- c("WALK")

max_trip_duration <- 60  # minutes

departure_datetime <- as.POSIXct("01-03-2021 14:00:00", 
                                 format = "%d-%m-%Y %H:%M:%S",
                                 tz = "Europe/London")

# estimating the travel time
TTM_Oxford_walk <- travel_time_matrix(r5r_core,
                          origins = points_o,
                          destinations = points_d,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_trip_duration = max_trip_duration)
```

Now for Driving 

```{r driving travel times}
# routing paramaters
mode <- c("CAR")

max_trip_duration <- 60  # minutes

departure_datetime <- as.POSIXct("01-03-2021 14:00:00", 
                                 format = "%d-%m-%Y %H:%M:%S",
                                 tz = "Europe/London")

# estimating the travel time
TTM_Oxford_drive <- travel_time_matrix(r5r_core,
                          origins = points_o,
                          destinations = points_d,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_trip_duration = max_trip_duration)
```

We now have three modes of transport (measure of access) to food banks in the area - public transport, walking, and driving. As you can see many more LSOAs have access to a food bank centre within a one hour drive, compared to one hour walking or public transport.

### Getting minimum travel time

#### Walking

Let's start with walking. This removes origin-destination pairs that have two or more values for travel time under one hour by taking the minimum value.

```{r as some pairs have more than on value, lets select the minimum value}
#select minimum value between pairs 
TTM_Oxford_walk2 <- TTM_Oxford_walk %>%
  group_by(to_id, from_id) %>% 
  slice_min(travel_time_p50, n = 1, with_ties = FALSE, na_rm = FALSE) %>%
  ungroup()

```

Next we take the long tibble and make it wide. The number of rows has decreased as not all LSOAs will have accessibility to a centre within one hour.

```{r matrix of min times between pairs}
#pivot wider
TTM_Oxford_walk3 <- TTM_Oxford_walk2 %>%
  pivot_wider(names_from = to_id, values_from = travel_time_p50)

head(TTM_Oxford_walk3)
```

#### Driving

Let's repeat this for driving and public transport.

```{r cleaning driving data}
#select minimum value between pairs 
TTM_Oxford_drive2 <- TTM_Oxford_drive %>%
  group_by(to_id, from_id) %>% 
  slice(which.min(travel_time_p50))

#check travel times as numeric
TTM_Oxford_drive2$travel_time_p50 <- as.numeric(TTM_Oxford_drive2$travel_time_p50)

#pivot wider to create a matrix
TTM_Oxford_drive3 <- TTM_Oxford_drive2 %>%
  pivot_wider(names_from = to_id, values_from = travel_time_p50)

head(TTM_Oxford_drive3)
```

#### Public transport

```{r cleaning public transport data}
#select minimum value between pairs 
TTM_Oxford_PT2 <- TTM_Oxford_PT %>%
  group_by(to_id, from_id) %>% 
  slice(which.min(travel_time_p50))

#check travel times as numeric
TTM_Oxford_PT2$travel_time_p50 <- as.numeric(TTM_Oxford_PT2$travel_time_p50)

#pivot wider to create a matrix
TTM_Oxford_PT3 <- TTM_Oxford_PT2 %>%
  pivot_wider(names_from = to_id, values_from = travel_time_p50)

head(TTM_Oxford_PT3)
```

### Join to LSOAs

Now we have the accessibility data in a usable form that we can join to our Oxford LSOAs data frame, we can calculate mean and minimum travel times for each LSOA for each mode of transport. R5R might find more than 1 food bank that is reachable for each centroid, so we can identify the minimum (and mean) travel time from the origin centroid.

```{r mean and min time from all OD pairs}

TTM_Oxford_drive3 <- TTM_Oxford_drive3 %>%
  # forces operation to happen row by row - find the smallest time in the row (not column)
  rowwise() %>%
  mutate(
    minimum_time_drive = min(c_across(2:14), na.rm = TRUE),
    mean_time_drive    = mean(c_across(2:14), na.rm = TRUE)
  ) %>%
  ungroup()


TTM_Oxford_walk3 <- TTM_Oxford_walk3 %>%
  rowwise() %>%
  mutate(
    minimum_time_walk = min(c_across(3:15), na.rm = TRUE),
    mean_time_walk  = mean(c_across(3:15), na.rm = TRUE)
  ) %>%
  ungroup()

TTM_Oxford_PT3 <- TTM_Oxford_PT3 %>%
  rowwise() %>%
  mutate(
    minimum_time_PT = min(c_across(2:14), na.rm = TRUE),
    mean_time_PT    = mean(c_across(2:14), na.rm = TRUE)
  ) %>%
  ungroup()


```

Now we need to join this to the Oxford LSOAs dataframe and map it.

```{r rename column and select columns needed}
#rename to from id columns for joining purposes
#select only the ID column and mean and minimum times

TTM_Oxford_drive3 <- TTM_Oxford_drive3 %>%
  select(from_id, mean_time_drive, minimum_time_drive) %>%
  rename(., lsoa21cd = "from_id")

TTM_Oxford_walk3 <- TTM_Oxford_walk3 %>%
  select(from_id, mean_time_walk, minimum_time_walk) %>%
  rename(., lsoa21cd = "from_id")

TTM_Oxford_PT3 <- TTM_Oxford_PT3 %>%
  select(from_id, mean_time_PT, minimum_time_PT) %>%
  rename(., lsoa21cd = "from_id")

Oxford_LSOAs <- Oxford_LSOAs %>%
  left_join(.,
            TTM_Oxford_drive3,
            by = "lsoa21cd") %>%
  left_join(.,
            TTM_Oxford_walk3,
            by = "lsoa21cd") %>%
  left_join(.,
            TTM_Oxford_PT3,
            by = "lsoa21cd")
```

remove any inf values as these won't map

```{r}
Oxford_LSOAs <- Oxford_LSOAs %>%
  mutate(
    minimum_time_walk  = ifelse(is.infinite(minimum_time_walk), NA, minimum_time_walk),
    minimum_time_drive = ifelse(is.infinite(minimum_time_drive), NA, minimum_time_drive),
    minimum_time_PT    = ifelse(is.infinite(minimum_time_PT), NA, minimum_time_PT)
  )
```

Map

```{r map for min PT travel time, include=TRUE}
#map for minimum public transport time
Travel_time_map <-
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "minimum_time_PT", title = "Minimum time to food bank (mins)", style = "cont", n=10, palette = "-Spectral",
          colorNA = "white", # colour of missing data
          textNA = "Public transport beyond one hour") +
    tm_borders(col = "grey", lwd = 0.01) +
  tm_shape(Oxford_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_City_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) +
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "PT accessibility in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) 

Travel_time_map 
```

```{r map for min driving time, include=TRUE}
#map for minimum driving time
Travel_time_map <-
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "minimum_time_drive", title = "Minimum time to food bank (mins)", style = "cont", n=10, palette = "-Spectral",
          colorNA = "white", # colour of missing data
          textNA = "Driving beyond one hour") +
    tm_borders(col = "grey", lwd = 0.01) +
  tm_shape(Oxford_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_City_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) +
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Driving accessibility in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) 

Travel_time_map 
```

```{r map for min walking time, include=TRUE}
#map for minimum walking time
Travel_time_map <-
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "minimum_time_walk", title = "Minimum time to food bank (mins)", style = "cont", n=10, palette = "-Spectral",
          colorNA = "white", # colour of missing data
          textNA = "Walking beyond one hour") +
    tm_borders(col = "grey", lwd = 0.01) +
  tm_shape(Oxford_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_City_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) +
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Walking accessibility in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) 

Travel_time_map 
```

From these three maps it is very clear that areas have unequal access to food bank centres, especially in the rural parts of Oxfordshire. However, do people in these areas of low accessibility need greater access to centres.

## Examining demand

In this last section we will examine two determinants of food bank use. Let's take housing tenure (socially rented homes) and disability. This is census data from 2021 at the LSOA scale.

```{r housing tenure}
#Type of tenure of housing
Tenure <- read_csv("prac5_data/Tenure.csv") %>%
  clean_names() %>%
  # separate a single column into 2 by using : as the divider
  separate(lsoa21cd, c('lsoa_code', 'lsoa_name'), sep = ':') %>%
  #trimws = remove white space before and after
  mutate(lsoa_name = trimws(lsoa_name)) %>%
  mutate(lsoa_code = trimws(lsoa_code)) 

#select socially rented variable
Tenure <- Tenure %>%
  select(lsoa_code, lsoa_name, social_rented)

#households that are socially rented
Tenure$social_rented <- as.numeric(Tenure$social_rented)
favstats(Tenure$social_rented)

#social rented
ggplot(Tenure,
       aes(x=social_rented)) + 
  geom_histogram(aes(y=after_stat(density)), color="black", fill="white") +
  geom_vline(aes(xintercept=mean(social_rented)),
            color="blue", linetype="dashed", size=1) +
 geom_density(alpha=.2, fill="#FF6666")
```

```{r disability}
#Disability under the Equality Act
Disability <- read_csv("prac5_data/Disability.csv") %>%
  clean_names() %>%
  separate(lsoa, c('lsoa_code', 'lsoa_name'), sep = ':') %>%
  mutate(lsoa_name = trimws(lsoa_name)) %>%
  mutate(lsoa_code = trimws(lsoa_code)) 

#select socially rented variable
Disability <- Disability %>%
  select(lsoa_code, lsoa_name, disabled_under_the_equality_act)

#households with 1+ disabled person
Disability$disabled_under_the_equality_act <- as.numeric(Disability$disabled_under_the_equality_act)
favstats(Disability$disabled_under_the_equality_act)

#disability
ggplot(Disability,
       aes(x=disabled_under_the_equality_act)) + 
  geom_histogram(aes(y=after_stat(density)), color="black", fill="white") +
  geom_vline(aes(xintercept=mean(disabled_under_the_equality_act)),
            color="blue", linetype="dashed", size=1) +
 geom_density(alpha=.2, fill="#FF6666") 
```

::: callout-note
**Question:**

What can we say about the distribution of these two *demand* variables? And how could we customise these histograms to improve the look of them?
:::

```{r join two demand variables to spatial data for LSOAs}
#rename lsoa code variable
Disability <- Disability %>%
  rename(., "lsoa21cd" = "lsoa_code") %>%
  rename(., "lsoa21nm" = "lsoa_name")

Tenure <- Tenure %>%
  rename(., "lsoa21cd" = "lsoa_code") %>%
  rename(., "lsoa21nm" = "lsoa_name")

#join to Oxford LSOA data frame
Oxford_LSOAs <- Oxford_LSOAs %>%
  left_join(.,
            Disability, 
            by = "lsoa21cd") %>%
  left_join(.,
            Tenure, 
            by = "lsoa21cd")
```

Let's make a map of disability (% of households with one or more disabled person living there) and socially rented homes.

```{r map for disability, echo=FALSE, include=TRUE}

Disability_map <-
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "disabled_under_the_equality_act", title = "Disability (% of HHs)", style = "cont", n=10, palette = "-Spectral",
          colorNA = "white", # colour of missing data
          textNA = "Walking beyond one hour") +
    tm_borders(col = "grey", lwd = 0.01) +
  tm_shape(Oxford_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_City_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) +
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Disability in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) 

Disability_map
```

```{r map for socially rented homes, echo=FALSE, include=TRUE}

Social_rent_map <-
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "social_rented", title = "Socially rented (% of HHs)", style = "cont", n=10, palette = "-Spectral") +
    tm_borders(col = "grey", lwd = 0.01) +
  tm_shape(Oxford_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_City_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) +
  tm_layout(legend.outside = TRUE, frame = FALSE, 
            main.title = "Social renting in & around Oxford", 
            main.title.position = "centre",
            main.title.size = 0.9) 

Social_rent_map
```

### Identifying under-served areas

Firstly, what are the national averages for disability and social renting across England and Wales?

```{r}
mean(Tenure$social_rented, na.rm = TRUE)

mean(Disability$disabled_under_the_equality_act, na.rm = TRUE)
```

-   Socially renting - 16.9% of households
-   Disability - 17.7% of households

And what are the Oxford and surrounding area averages?

```{r}
mean(Oxford_LSOAs$social_rented, na.rm = TRUE)

mean(Oxford_LSOAs$disabled_under_the_equality_act, na.rm = TRUE)
```

-   Socially renting - 14.5% of households
-   Disability - 14.6% of households

They are lower than the national averages.

Let's take a look for the three travel times too.

```{r}
mean(Oxford_LSOAs$mean_time_drive, na.rm = TRUE)

mean(Oxford_LSOAs$mean_time_walk, na.rm = TRUE)

mean(Oxford_LSOAs$mean_time_PT, na.rm = TRUE)
```

Very simply, if we take the areas that are below the area average for social renting and disability and one of the accessibility measures, which places in and around Oxford are under served to food banks based on high demand and lower accessibility?

```{r areas that are underserved}
#set some parameters to identify which areas are under served
Oxford_LSOAs <- Oxford_LSOAs %>%
  mutate(Underserved_areas = case_when
         (social_rented >= 14.5 & disabled_under_the_equality_act >= 14.6 & mean_time_PT >= 32 ~ "Underserved", 
        TRUE ~ "Served"))
```

```{r creating a map of under served areas}
#creating a buffer around the LSOAs
buffer <- Oxford_LSOAs %>%
  filter(Underserved_areas == "Underserved") %>%
  st_buffer(., 500) %>%
  st_union()

 Underserved_map <-
  tm_shape(Oxford_LSOAs) +
  tm_fill(col = "Underserved_areas", style = "cat", title = "Underserved Areas", palette = "Blues") +
  tm_borders(col = "grey", lwd = 0.1) +
  tm_shape(buffer) +
  tm_polygons(alpha = 0.1) +
  tm_shape(Oxford_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_shape(Oxford_City_LAs) +
    tm_borders(col = "black", lwd = 1) + 
  tm_scale_bar(position = c("right", "top")) + 
  tm_compass(position = c("right", "top")) +
  tm_layout(main.title = "Under-served areas to food banks", main.title.position = "centre",
            legend.outside = TRUE, frame = FALSE, main.title.size = 1) 

Underserved_map
```

## Quick summary

We can use accessibility tools and census data to assess where in an area is under-served to particular services.

-   How would we improve this analysis?
-   What are the limitations of the data used?
-   How could we improve the visualisations?
